Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-13 manuscript/manuscript.tex
more theory section stuff

When mutations are weakly deleterious, their allele frequencies are strongly
influenced by stochastic perturbations and their frequency trajectories are no
longer well-approximated by deterministic models. When genetic drift is the
only source of randomness in allele frequency change, the relevant scale of
stochastic perturbations is determined by the drift-effective population size
$N_d$ (e.g. due to only non-selective process). For reasons that will soon be
clear, we consider a general expression of the drift-effective population size
that allows for extra non-heritable variability in offspring number $V_k$,
%
\begin{align}
    N_d = \frac{2N}{V_k + 2}
\end{align}

where if $V_k = 2$, this reduces to the classic Wright--Fisher model
\parencite{Wright1938-tv}.

@@

To further complicate matters, both strong and weakly deleterious mutations
create \emph{heritable} fitness variation. In turn, the presence of heritable
fitness variation generates another source of randomness in allele frequency
change known as genetic \emph{draft} \parencite{Neher2013-dz}. Genetic draft
occurs when alleles become randomly associated with different fitness
backgrounds each generation, which stochastically perturbs their trajectories.
Like genetic drift, perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, stochastic
perturbations caused by draft are autocorrelated across generations as long an
allele does not recombine or segregate off its fitness background and the
selective environment does not change
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. When selection is
strong, draft can generate multiple-merger coalescences that create
discontinuities in allele frequency trajectories, making diffusion
approximations ill-suited \parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}.
However, when selection is weak and fitness is polygenic, the stochasticity
introduced by draft can be approximated by assuming draft just increases the
rate of drift. This is often expressed as rescaling the effective population
size

Though there are subtle differences in the effective population sizes

>>>>

When mutations are weakly deleterious, their allele frequencies are strongly
influenced by stochastic perturbations and their frequency trajectories are no
longer well-approximated by deterministic models. When genetic drift is the
only source of randomness in allele frequency change, the relevant scale of
stochastic perturbations is determined by the drift-effective population size
$N_d$ (e.g. due to only non-selective process). For reasons that will soon be
clear, we consider a general expression of the drift-effective population size
that allows for extra non-heritable variability in offspring number $V^2$,
%
\begin{align}
    N_d = \frac{2N}{V^2 + 2}
\end{align}

where if $V^2 = 2$, this reduces to the classic Wright--Fisher model
\parencite{Wright1938-tv}.

@@

To further complicate matters, both strong and weakly deleterious mutations
create \emph{heritable} fitness variation. In turn, the presence of heritable
fitness variation generates another source of randomness in allele frequency
change known as genetic \emph{draft} \parencite{Neher2013-dz}. Genetic draft
occurs when alleles become randomly associated with different fitness
backgrounds each generation, which stochastically perturbs their trajectories.
Like genetic drift, perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, stochastic
perturbations caused by draft are autocorrelated across generations as long an
allele does not recombine or segregate off its fitness background and the
selective environment does not change
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. When selection is
strong, draft can generate multiple-merger coalescences that create
discontinuities in allele frequency trajectories, making diffusion
approximations ill-suited \parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}.
However, when selection is weak and polygenic, draft can be approximated as an
increase in the rate of drift. The key insight is that the draft rescaling
factor is like extra variability in offspring number, but because of
autocorrelation due to linkage and similar environments, the heritable fitness
variation $C^2$ becomes inflated by a factor $Q^2$,

\begin{align}
    N_e = \frac{N}{C^2 Q^2 + 1}.
\end{align}

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof).

Note that assuming $V^2=2$.

are subtle differences in the effective population sizes

This is often
expressed as rescaling the effective population size by
